Human capital mediates natural selection in contemporary humans

### 2021-08-10 model-appendix.Rmd
Final final tweaks including new end of intro

We can now gather the facts stated in the main text.

\textbf{Fact 1}: for $\sigma<1$, total fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$
is decreasing in human capital $h$.

@@

\textbf{Fact 2 part 1}: the fertility-human capital relationship is
closer to 0 at high levels of $h$.

@@

\textbf{Fact 2 part 2}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among higher earners.

\textbf{Fact 4}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker at high levels of education.

@@

\textbf{Fact 5}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among those who start fertility late.

Lastly, we prove fact 3. Differentiating $dN_{t}^{*}/dh$ in (\ref{eq:dNdh1})
with respect to $b$, for when $N_{1}^{*}>0$ gives:
\begin{align*}
\frac{d^{2}N_{t}^{*}}{dhdb} & =\frac{2\sigma-2}{2\sigma-1}b^{(3-4\sigma)/(2\sigma-1)}\left(\frac{1}{a}\right)^{1/(2\sigma-1)}\frac{1-\sigma}{2\sigma-1}h^{(\sigma-1)^{2}/(\sigma(2\sigma-1))}
\end{align*}

@@

\textbf{Fact 3}: for $\sigma<1$, the fertility-human capital relationship
is more negative when the burden of childcare $b$ is larger.

>>>>

We can now gather the theoretical predictions stated in Table \ref{tab:theory}.

\textbf{Prediction 1}: for $\sigma<1$, total fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$
is decreasing in human capital $h$.

@@

\textbf{Prediction 2 part 1}: the fertility-human capital relationship is
closer to 0 at high levels of $h$.

@@

\textbf{Prediction 2 part 2}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among higher earners.

\textbf{Prediction 4}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker at high levels of education.

@@

\textbf{Prediction 5}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among those who start fertility late.

Lastly, we prove prediction 3. Differentiating $dN_{t}^{*}/dh$ in (\ref{eq:dNdh1})
with respect to $b$, for when $N_{1}^{*}>0$ gives:
\begin{align*}
\frac{d^{2}N_{t}^{*}}{dhdb} & =\frac{2\sigma-2}{2\sigma-1}b^{(3-4\sigma)/(2\sigma-1)}\left(\frac{1}{a}\right)^{1/(2\sigma-1)}\frac{1-\sigma}{2\sigma-1}h^{(\sigma-1)^{2}/(\sigma(2\sigma-1))}
\end{align*}

@@

\textbf{Prediction 3}: for $\sigma<1$, the fertility-human capital relationship
is more negative when the burden of childcare $b$ is larger.

### 2021-08-10 why-natural-selection.Rmd
Final final tweaks including new end of intro

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents and
younger parents, while among couples and older parents (22+) effects are more
evenly balanced.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

@@

Suppose that certain genetic variants correlate with human capital -- skills or 
other characteristics that affect an individual's earnings in the labour market.
These variants may then be associated with opposing effects on fertility. The income
effect will lead to natural selection in favour of earnings-increasing variants
(or variants that are merely associated with higher earnings). The substitution
effect will do the reverse.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
and in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
predictions, along with corresponding empirical results for the correlation between
polygenic scores and fertility.[^one-period-results]

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. This need not imply
that a given polygenic score is causal of either fertility or human capital. Instead
scores may simply correlate with causally relevant environments, for example via
"genetic nurture" effects [@kong2018nature] or population stratification. To
test whether polygenic scores cause fertility, we estimate equation
\@ref(eq:regression) among full siblings, controlling for a full set of sibling
group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
sample size, all within-sibling effects are insignificant after Bonferroni
correction. However, effect sizes are positively correlated with effect sizes
from the pooled model, and about 1/3 smaller (regressing within-sibling on
pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
with the ~50% decrease in heritability in within-sibling GWASs on age at first
birth and educational attainment [@Howe2021.03.05.433935]. We see these results
as providing weak evidence that polygenic scores cause fertility, with effects
being at least partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores (EA2 and EA3), effect sizes are reduced by about 20%.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our results
suggest that the substitution effect dominates for single parents and younger
parents, while among couples and older parents (22+) effects are more evenly
balanced. Thus, contemporary natural selection on polygenic scores can be
explained by their correlation with earnings-increasing human capital. Below, we 
show that a simple model of human capital, education and fertility choices 
can give rise to our key empirical results.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

@@

Suppose that certain genetic variants correlate with human capital -- skills or
other characteristics that affect an individual's earnings in the labour market
[@mincer1958investment; @becker1964human]. These variants may then be associated
with opposing effects on fertility. The income effect will lead to natural
selection in favour of earnings-increasing variants (or variants that are merely
associated with higher earnings). The substitution effect will do the reverse.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$
in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
predictions, along with corresponding empirical results for the correlation between
polygenic scores and fertility.[^one-period-results]

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. This need not imply
that a given polygenic score is causal of either fertility or human capital. Instead
scores may simply correlate with causally relevant environments, for example via
"genetic nurture" effects [@kong2018nature] or population stratification. To 
investigate the presence of these indirect genetic effects, we estimate equation
\@ref(eq:regression) among full siblings, controlling for a full set of sibling
group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
sample size, all within-sibling effects are insignificant after Bonferroni
correction. However, effect sizes are positively correlated with effect sizes
from the pooled model, and about 1/3 smaller (regressing within-sibling on
pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
with the ~50% decrease in heritability in within-sibling GWASs on age at first
birth and educational attainment [@Howe2021.03.05.433935]. We see these results
as providing weak evidence that polygenic scores cause fertility, with effects
being at least partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores, effect sizes are reduced by about 20%.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects. Respondents weighted by Age/Qualification.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )
